Relationships in the extremes and their influence on competition and coexistence

### 2021-06-24 paper/Paper.Rnw

For a general model, it is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$. For simplicity, we here assume 
the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though this assumption can be circumvented, we make the assumption because it simplifies the
presentation without obscuring main ideas. 
For the lottery model, we take $E_i(t)=B_i(t)$, henceforth assuming that fecundity depends strictly on
the environment. Competition, $C_i(t)$, in the lottery model is the number of new offspring
divided by the number of open sites, $C_i(t)=( B_1(t)N_1(t)+B_2(t)N_2(t) )/\delta N$, which
does not depend on $i$. Competition in the lottery model occurs at the stage of juveniles occupying open sites. 
Species growth rates for the lottery model can then be written 
$r_i(t)=\ln[1-\delta+E_i(t)/C_i(t)]=\ln[1-\delta+\delta N B_i(t)/(B_1(t)N_1(t)+B_2(t)N_2(t))]$.

In general, the storage effect for species $i$ is the contribution of covariation between $E_i(t)$ and 
$C_i(t)$ to a difference between invader and resident mean growth rates, when $i$ is the invader.
Specifically, storage effects for species $i$ are 
\begin{equation}
\Delta I_i = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ refers to the mean ``invasion growth rate'' of species $i$, that is, the growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ refers to the mean steady-state 
growth rate of species $j$ when species $i$ is absent (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative model scenario, identical except that
the bivariate random variable $(E_i,C_i)$ has been replaced by a random variable $(E_i^\shp,C_i^\shp)$ 
with the same marginal distributions but independent components. The so-called \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species, see \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the definition of $q_{ij}$, here 
omitted for brevity.

For the lottery model, it is straightforward to show that $\olr_{i \bs i}=\E[\ln(1-\delta+\delta B_i(t)/B_j(t))]$,
$\olrs_{i \bs i} = \E[\ln(1-\delta+\delta B_i^\shp(t)/B_j(t))]$, and $\olrs_{j \bs i} = \E[\ln(1-\delta+\delta B_j^\shp(t)/B_j(t))]$, where $\E(\cdot)$ is expected value.
Here, $B_i^\shp$ is a random variable distributed in the same way as $B_i$, but independent of it; and
$B_j^\shp$ is distributed the same as $B_j$, but is independent of it. We show in SI section \ref{SI-SIsect:scaling_factors}
that $q_{ij}=\overline{B}_i/((1-\delta)\overline{B}_j+\delta \overline{B}_i)$, where, as above, the overbar 
denotes expected value. 
%***DAN: Note this might be wrong. Some of the language in Ellner et al (2016) indicates q_{ij} here should be 1,
%even if \mu_1 and \mu_2 are distinct, as long as we use the same \delta for both species (which we do).
%But when I try to follow the definitions of the same paper to compute q_{ij} I get the quantity listed above,
%which is not 1. I'll need to get to the bottom of this, possibly by speaking to Ellner, and then put the right 
%value here, and also write the corresponding sup mat section accordingly (though you may not need that section, 
%it may be enough to just cite Ellner and Chesson if you go with q_ij=1.
Since we use log-normally distributed $B_i$, 
$\overline{B}_i=\exp(\mu_i+\sigma^2/2)$, so $q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$.
%***DAN: This statement also needs adjusting if you go with q_ij=1
So $\Delta I_i$ for the lottery model can be computed by using simulations \citep{Ellner_2016}
to compute arbitrarily good approximations for $\olr_{i \bs i}$, $\olrs_{i \bs i}$,
and $\olrs_{j \bs i}$, and then plugging them and the value for $q_{ij}$ into (\ref{eq:storage_effects_gen}).

Having just reviewed storage effects, we now specify how to decompose storage effects into
contrbutions due to asymmetric tail associations and contributions due
to $EC$ covariance \emph{per se}. The storage effect for species $i$ was computed by comparing
a difference between invader and resident mean growth rates for the original model to the same quantity calculated for a 
modified model for which $E_i$ and $C_i$ were rendered independent (see (\ref{eq:storage_effects_gen})). 
Because $E_i$ and $C_i$ were rendered 
completely independent in the modified model, storage effects quantify the contribution of the 
whole relationship between $E_i$ and $C_i$, in all its aspects, to the invader/resident growth rate
difference. However, we can instead consider a distribution $(E_i^\nat,C_i^\nat)$ with the same
marginal distributions as $(E_i,C_i)$ and with some but not all aspects of the relationship 
between the components modified relative to $(E_i,C_i)$; specifically, tail associations are rendered symmetric
(see below for how this is done). We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_i^\nat)$,
and we consider the decomposition 
\begin{align}
\Delta I_i &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \\
&= I_i^{(\text{ata})} + I_i^{(\text{cps})}.\label{eq:storage_effects_further}
\end{align}
The superscript ``ata" stands for ``asymmetric tail associations'' because this component of storage effects is due to 
asymmetric tail associations between $E_i$ and $C_i$. The superscript ``cps" stands for ``covariance
\emph{per se}'' because this component of storage effects is due to $EC$ covariance \emph{per se}.

@@

>>>>

For a general model, it is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$. For simplicity, we here assume 
the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though this assumption can be circumvented, we make the assumption because it simplifies the
presentation without obscuring ideas. 
For the lottery model, we take $E_i(t)=B_i(t)$, henceforth assuming that fecundity depends strictly on
the environment. Competition, $C_i(t)$, in the lottery model is the number of new offspring
divided by the number of open sites, $C_i(t)=( B_1(t)N_1(t)+B_2(t)N_2(t) )/\delta N$, which
does not depend on $i$. Competition in the lottery model occurs at the stage of juveniles occupying open sites. 
Species growth rates for the lottery model can then be written 
$r_i(t)=\ln[1-\delta+\delta N B_i(t)/(B_1(t)N_1(t)+B_2(t)N_2(t))]=\ln[1-\delta+E_i(t)/C_i(t)]$.

In general, the storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between invader and resident mean growth rates, when $i$ is the invader.
Specifically, storage effects for species $i$ are 
\begin{equation}
\Delta I_i = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ refers to the mean ``invasion growth rate'' of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ refers to the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative model scenario, identical except that
influences of environment and competition on the invader or resident have been rendered independent.  
A subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species, see \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the definition of $q_{ij}$.

For the lottery model, when $i$ is an invader, the competition experienced by either 
species is obtained from the expression for $C_i=C_j$ by setting $N_i=0$ and $N_j=N$; and likewise to get
$\olr_{i \bs i}$ from the expression for $r_i$ for the model. Thus 
$C_{i \bs i}(t)=C_{j \bs i}(t)=B_j(t)/\delta=E_j(t)/\delta$, and 
$\olr_{i \bs i} = \E[\ln(1-\delta+\delta B_i/B_j)]=\E[\ln(1-\delta+E_i /C_{i \bs i})]$. 
Here $\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions for which
an overbar would be cumbersome. Likewise, 
$\olr_{j \bs i} = \E[\ln(1-\delta+\delta B_j/B_j)]=\E[\ln(1-\delta+E_j/C_{j \bs i})]$,
though we already knew this equals $0$. An expression for $\olrs_{i \bs i}$ is then obtained from the 
above expression for $\olr_{i \bs i}$ by replacing the bivariate random variable $(E_i,C_{i \bs i})=(B_i,B_j/\delta)$
by a random variable with the same marginals, but independent components, $(B_i^\shp,B_j^\shp /\delta)$.
Thus $\olrs_{i \bs i} = \E[\ln(1-\delta+\delta B_i^\shp/B_j^\shp)]$. Likewise, a formula for 
$\olrs_{j \bs i}$ is available by replacing $(E_j,C_{j \bs i})$ in an analogous way. We use 
$(E_j^\shp,C_{j \bs i}^\shp)=(B_j^\shp, B_j /\delta)$, so 
$\olrs_{j \bs i}=\E[\ln(1-\delta+\delta B_j^\shp / B_j)]$.
We show in SI section \ref{SI-SIsect:scaling_factors}
that $q_{ij}=\overline{B}_i/((1-\delta)\overline{B}_j+\delta \overline{B}_i)$. 
%***DAN: Note this might be wrong. Some of the language in Ellner et al (2016) indicates q_{ij} here should be 1,
%even if \mu_1 and \mu_2 are distinct, as long as we use the same \delta for both species (which we do).
%But when I try to follow the definitions of the same paper to compute q_{ij} I get the quantity listed above,
%which is not 1. I'll need to get to the bottom of this, possibly by speaking to Ellner, and then put the right 
%value here, and also write the corresponding sup mat section accordingly (though you may not need that section, 
%it may be enough to just cite Ellner and Chesson if you go with q_ij=1.
Since we use log-normally distributed $B_i$, 
$\overline{B}_i=\exp(\mu_i+\sigma^2/2)$, so $q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$.
%***DAN: This statement also needs adjusting if you go with q_ij=1
So $\Delta I_i$ for the lottery model can be computed by using simulations \citep{Ellner_2016}
to very accurately approximate $\olr_{i \bs i}$, $\olrs_{i \bs i}$,
and $\olrs_{j \bs i}$, and then plugging them and $q_{ij}$ into (\ref{eq:storage_effects_gen}).

Having just reviewed storage effects, we now specify how to decompose storage effects into
contributions due to asymmetric tail associations and due
to $EC$ correlation \emph{per se}. 
Storage effects for species $i$ were computed by comparing
differences between invader and resident mean growth rates for the original model and for a 
modified model for which the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
were rendered independent of each other [see (\ref{eq:storage_effects_gen})]. 
Because these components were rendered 
completely independent in the modified model, storage effects quantify the contribution of the 
whole $EC$ relationship, in all its aspects, to the invader/resident growth rate
difference. But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between the components altered relative to the original distributions. Specifically, asymmetries of tail association are removed -- see below for how this is done. We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta I_i &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \\
&= I_i^{(\text{ata})} + I_i^{(\text{cps})}.\label{eq:storage_effects_further}
\end{align}
The superscript ``ata" stands for ``asymmetric tail associations'' because this component of storage effects is due to 
asymmetries of tail association between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$. 
The superscript ``cps" stands for ``correlation
\emph{per se}'' because this component is due to $EC$ correlation \emph{per se}, after 
removing the effects of asymmetric tail associations.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components.

How are asymmetries of tail association removed while leaving $E$ and $C$ marginal distributions and overall 
correlation unchanged? We assume that the cummulative distribution functions (cdfs) of $E_i$ and 
$C_{i \bs i}$, which we denote $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively, are strictly monotonic and continuous. 
Letting $\varphi$ be the cdf of a standard normal distribution, we consider the random variables
$\varphi^{-1} \circ F_{E_i} (E_i)$ and $\varphi^{-1} \circ F_{C_{i \bs i}}(C_{i \bs i})$, which are standard normally
distributed, although they are not necessarily jointly bivariate normal.  Let $\rho_i$ be the covariance of
these random variables. We then define a bivariate normal random variable $(e_i,c_{i \bs i})$ such that
$\var(e_i)=\var(c_{i \bs i})=1$, and denote $\cov(e_i,c_{i \bs i})$ by $\tilde{\rho}_i$. We define 
$E_i^\nat = F_{E_i}^{-1} \circ \varphi(e_i)$
and $C_{i \bs i}^\nat=F_{C_{i \bs i}}^{-1} \circ \varphi(c_{i \bs i})$. The value of $\tilde{\rho}_i$ is then determined 
in one of 
four alternative ways: $\tilde{\rho}_i = \rho_i$; or so that the Pearson, Spearman or Kendall correlation of 
$E_i^\nat$ and $C_{i \bs i}^\nat$ matches that of $E_i$ and $C_{i \bs i}$. The choice of bivariate normal 
$(e_i,c_{i \bs i})$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are symmetrically tail associated in the sense
described in the Introduction.
%***DAN: This means you have to introduce this well, and explain why the way we define tail assoc is
%the *right* way
For those familar with the language of copula statistics (see, e.g., \citealp{Ghosh_2020_AER} and \citealp{Genest_2007}), 
we replaced the copula of $(E_i,C_{i \bs i})$ with a normal copula to form $(E_i^\nat,C_{i \bs i}^\nat)$.
The choice of $\tilde{\rho}_i$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are correlated to the same extent as 
$E_i$ and $C_{i \bs i}$, according to whichever definition of correlation is considered preferable. Thus $E_i^\nat$ and 
$C_{i \bs i}^\nat$ have had their tail association rendered symmetric while retaining their correlation \emph{per se},
as desired. The random variable $(E_j^\nat, C_{j \bs i}^\nat)$ is created in an analogous way.

For the lottery model, recall that $E_i=B_i$ and $C_{i \bs i} = B_j/\delta$. So 
$\varphi^{-1} \circ F_{E_i}(E_i) = \frac{b_i-\mu_i}{\sigma}$ and 
$\varphi^{-1} \circ F_{C_{i \bs i}}(C_{i \bs i})=\frac{b_j - \mu_j}{\sigma}$.  
Also, $\rho_i = \cov(\frac{b_i-\mu_i}{\sigma},\frac{b_j-\mu_j}{\sigma})$, which equals the 
Pearson correlation of $b_i$ and $b_j$. For the lottery model, it seems natural to make the choice
$\tilde{\rho}_i=\rho_i$, which straightforwardly implies that $(E_i^\nat,C_{i \bs i}^\nat)$
can be written as $(\exp(\tilde{b}_i),\exp(\tilde{b}_j))$, where $(\tilde{b}_i,\tilde{b}_j)$
is bivariate normally distributed with mean $(\mu_i,\mu_j)$, $\var(\tilde{b}_i)=\var(\tilde{b}_j)=\sigma$,
and $\cov(\tilde{b}_i,\tilde{b}_j)=\cov(b_i,b_j)$. So $(\tilde{b}_i,\tilde{b}_j)$ has the same
marginal distributions and covariance matrix as $(b_i,b_j)$, but it is bivariate normal, whereas
$(b_i,b_j)$ itself need not be. Given these definitions, it is straightforward to compute the 
components of (\ref{eq:storage_effects_further}) for the lottery model using the simulation 
approach of \cite{Ellner_2016,Ellner_2019}, though in the next section we describe some tricks
that make these computations efficient.

The other mechanism in modern coexistence theory that apply in our modelling context (i.e., purely temporal variation)
make no reference to covariation between quantities, whereas asymmetries of tail associations are 
defined in a covariation context. Therefore, storage effects are the only mechanism of coexistence
for which asymmetries of tail association are relevant. Details are given in SI section 
\ref{SIsect:other_mechanisms}, and we return briefly to related ideas in the Discussion.
%***DAN: In the Discussion, I intend to talk about how possible changes in tail association 
%patterns of env vars in nature may influence coexistence of species. It won't influence the 
%other mechanisms, though changes in the extreme values of environmental variables, known
%to be occurring, might.

@@

%***DAN: Next steps:
%Write to Ellner asking question about scaling factors
%Decide what we ought to be computing 
%Write Methods correspondingly
%Talk to Jasmin about it - did I miss anything that we already computed?
%Then work with Jasmin to get the code edited to compute that stuff
%Write results
%Once results have solidified, make a code suite version that is clean
